---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

*Introduction*

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

*Data*

The data for this assignment can be downloaded from the course web site.

The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

```{r}
echo = TRUE
setwd("~/Desktop/Coursera/Github/Repos/RepData_PeerAssessment1")
unzip("activity.zip", "activity.csv")
activity <- NULL
activity <- read.csv("activity.csv", header = T, sep = ",")
```

Created variables set to NULL.

```{r}
echo = TRUE
df_summary <- NULL
su2 <- NULL
su <- NULL
mn_int <- NULL
activity2 <- NULL
mean_su2 <- NULL
median_su2 <- NULL
activity2_weekend <- NULL
activity2_weekday <- NULL
mean_activity2_weekday <- NULL
mean_activity2_weekend <- NULL
```

## What is mean total number of steps taken per day?

Determining total number of steps per day.

```{r}
echo = TRUE
su <- tapply(activity$steps, activity$date, sum, na.rm=T)
```

Plotting histogram of the total number of steps taken each day.

```{r}
echo = TRUE
hist(su, xlab = "sum of steps per day",col="green",main = "histogram of steps per day")
```

Reporting the median total number of steps taken per day:
        
```{r}
echo = TRUE
mean_su <- round(mean(su))
median_su <- round(median(su))
print(c("The mean is",mean_su))
```

```{r}
print(c("The median is",median_su))
```

## What is the average daily activity pattern ?

A time series plot of the 5-minute interval and the average number of steps taken (averaged across all days) is shown below:
        
```{r}
echo = TRUE
mn_int <- tapply(activity$steps, activity$interval, mean, na.rm=T)
plot(mn_int ~ unique(activity$interval), type="l", xlab = "5-min interval", col = "purple")
```

Reporting the 5-minute interval (on average across all the days in the dataset) that contains the maximum number of steps: maximum number of steps and the value associated with it. 
        
```{r}
echo = TRUE
mn_int[which.max(mn_int)]
```

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing values may introduce bias into the data.

First, in order to visualize in which variable the NAs are:
        
```{r}
echo = TRUE
table(is.na(activity) == TRUE)
```

```{r}
summary(activity)
```

There are 2304 NA's.

*Strategy for filling in all of the missing values in the dataset*

For any NA is the step variable, the mean (of steps) of the corresponding interval is taken as the replacing value.

The 'mn_int' contains the mean for each single interval calculated over the 61 days. The right value coming from 'mn_int' is going to be used to replace the NA at the same interval.

```{r}
echo = TRUE
activity2 <- activity  
```

Creating dataset with no NAs

```{r}
for (i in 1:nrow(activity)){
        if(is.na(activity$steps[i])){
                activity2$steps[i]<- mn_int[[as.character(activity[i, "interval"])]]
        }
}
```

Plotting histogram of the total number of steps taken each day. The mean and median total number of steps taken per day are shown.

```{r}
echo = TRUE
su2 <- tapply(activity2$steps, activity2$date, sum, na.rm=T)
hist(su2, xlab = "sum of steps per day",col="green", main = "histogram of steps per day")
```

```{r}
mean_su2 <- round(mean(su2))
median_su2 <- round(median(su2))
```

The new values are :

```{r}        
echo = TRUE
print(c("The mean is",mean_su2))
```

```{r}
print(c("The median is",median_su2))
```

Comparing the new values with the “old” values:
        
```{r}
echo = TRUE
df_summary <- rbind(df_summary, data.frame(mean = c(mean_su, mean_su2), median = c(median_su, median_su2)))
rownames(df_summary) <- c("with NA's", "without NA's")
print(df_summary)
```

For comparison with NA's and without (see earlier):

```{r}
echo = TRUE
summary(activity2)
```

Confirmation that there are no more NAs in the steps variable.

## Are there differences in activity patterns between weekdays and weekends?

A new column is added to the dataframe, this column will contain the factor “weekday days”“ or "weekend days”.

```{r}
echo = TRUE
activity2$weekday <- c("weekday")
activity2[weekdays(as.Date(activity2[, 2])) %in% c("Saturday", "Sunday", "samedi", "dimanche", "saturday", "sunday", "Samedi", "Dimanche"), ][4] <- c("weekend")
table(activity2$weekday == "weekend")
```

```{r}
activity2$weekday <- factor(activity2$weekday)
```

In order to visualize the difference bewteen weekends and days of the week, a new dataframe is created. The lattice package is applied. 

Calculating the data

```{r}
echo = TRUE
activity2_weekend <- subset(activity2, activity2$weekday == "weekend")
activity2_weekday <- subset(activity2, activity2$weekday == "weekday")

mean_activity2_weekday <- tapply(activity2_weekday$steps, activity2_weekday$interval, mean)
mean_activity2_weekend <- tapply(activity2_weekend$steps, activity2_weekend$interval, mean)
```

Preparing the dataframe and the plot 

```{r}
echo = TRUE
library(lattice)
df_weekday <- NULL
df_weekend <- NULL
df_final <- NULL
df_weekday <- data.frame(interval = unique(activity2_weekday$interval), avg = as.numeric(mean_activity2_weekday), day = rep("weekday", length(mean_activity2_weekday)))
df_weekend <- data.frame(interval = unique(activity2_weekend$interval), avg = as.numeric(mean_activity2_weekend), day = rep("weekend", length(mean_activity2_weekend)))
df_final <- rbind(df_weekday, df_weekend)

xyplot(avg ~ interval | day, data = df_final, layout = c(1, 2), 
type = "l", ylab = "Number of steps")
```

From this plot, some differences in activity patterns between the average weekday and weekend are visible. it looks like the subject was more active earlier in the morning on weekdays. Activity continues until a later in the weekend.